---
title: "Timeline visualisation"
output:
  html_document:
    df_print: paged
---

Notebook demonstrating how to use the [timevis](https://github.com/daattali/timevis) R package using randomly generated events and time periods. 

Done to demonstrate how one may illustrate an individual's history. Specifically, inspired by this application in Health care: https://github.com/BenGlicksberg/PatientExploreR

```{r setup, message = FALSE}
# install.packages("timevis")
library(timevis)

library(data.table)
library(dtplyr)
library(dplyr, warn.conflicts = FALSE)
library(lubridate)
library(magrittr)
library(RColorBrewer)

# Parameters for data generation
num_ev <- 50
num_periods <- 10
start_date <- as.Date('2017-01-01')
end_date <- as.Date('2019-01-01')
# Parameters for styling and grouping
colour_col <- 'content_type'
group_col <- 'groups'
```

# Data generation

Generate events of different types (with no end date).

```{r event data generation}
types <- c("one", "two", "three")
events <- data.frame(ev_id = seq(1, num_ev, length = num_ev),
                     groups = "event",
                     subgroups = sample(types, num_ev, replace = TRUE),
                     start = sample(seq(start_date, end_date, by = "day"), num_ev),
                     end = as.Date(NA),
                     content = "event description",
                     stringsAsFactors = FALSE) %>%
  dtplyr::lazy_dt(.) %>%
  dplyr::mutate(content_type = subgroups) %>%
  data.table::as.data.table(.)
```

Generate time periods of different types. 

```{r time range generation}
period_types <- c("alpha", "beta", "gamma")
periods <- data.frame(ev_id = seq(num_ev, num_ev + num_periods, length = num_periods),
                      groups = "period",
                      subgroups = sample(period_types, num_periods, replace = TRUE),
                      start = sample(seq(start_date, end_date, by = "day"), num_periods),
                      end = sample(seq(start_date, end_date, by = "day"), num_periods),
                      content = "period description",
                     stringsAsFactors = FALSE) %>%
  dtplyr::lazy_dt(.) %>%
  dplyr::mutate(content_type = subgroups,
                end = as.Date(ifelse(end < start, start + 14, end), origin = "1970-01-01")) %>%
  data.table::as.data.table(.)
```

```{r combine events and periods}
all_data <- dplyr::bind_rows(events, periods)
```

# Colour style generation

Use RColorBrewer to create a qualitative palette of colours. This creates HEX codes, which then need to be pasted into a CSS style vector. 

```{r colour vector}
# Set group colours
n <- length(unique(all_data[[colour_col]]))
qual_col_pals <- RColorBrewer::brewer.pal.info[brewer.pal.info$category == 'qual',]
col_code <- sample(unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals))), n, replace = TRUE)
# Turn list of HEX codes into CSS styling string
col_vector <- paste("background-color:", col_code, "; border-color:", col_code)
# To also change the colour of the dots, it needs to be in a Shiny HTML tag: 
# https://stackoverflow.com/questions/57069911/how-do-i-change-the-color-of-the-points-in-the-timeline-visualization-using-time
```

```{r style dataframe}
# Create style based on colours above
styles <- data.frame(content = unique(all_data[[colour_col]]),
                     style_id = as.numeric(as.factor(unique(all_data[[colour_col]]))),
                     title = unique(all_data[[colour_col]]),
                     code = col_code,
                     style = col_vector)
```

# Group ID generation

Each group (named in the left margin) needs its own numeric ID. Timevis then requires a dataframe of group name and IDs to add the strings to the left margin. 

```{r group ID dataframe}
# Create group IDs
groups <- data.frame(content = unique(all_data[[group_col]]),
                     id = as.numeric(as.factor(unique(all_data[[group_col]]))),
                     title = unique(all_data[[group_col]]))
```

# Combine all data with style and group IDs

Merge the dataframe created above into the data, so we can prepare the final data as it is required by the timevis package. 

```{r add style and group IDs to main data}
prep_data <- copy(all_data)  %>%
  merge(., styles, how = 'left', by.x = colour_col, by.y = "content") %>%
  merge(., groups, how = 'left', by.x = group_col, by.y = "content") %>%
  .[, .(id = rownames(.),
        content = content,
        start = start,
        group = id,
        subgroup = subgroups,
        end = end,
        style = style)]
```

# Create actual timeline visualisation

Takes the prepared data (with the correct column names), and the groups dataframe to make sense of the group IDs. 

```{r visualise timeline}
timevis::timevis(prep_data, groups = groups, fit = TRUE, 
                 options = list(stack = FALSE))
```




